%-----------------------------------------------------------------------------
%
%               Template for LaTeX Class/Style File
%
% Name:         sigplanconf-template.tex
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------


\documentclass[natbib,10pt]{sigplanconf}

\usepackage{amsmath}
\usepackage{url}
\begin{document}

\conferenceinfo{ILC '09}{March 22--25, 2009, Cambridge, Massachusetts, USA.}
\copyrightyear{2009}
\copyrightdata{[to be supplied]}
\toappear{}

\titlebanner{DRAFT}        % These are ignored unless
\preprintfooter{A tutorial on using monads in Clojure and scheme}   % 'preprint' option specified.

\title{Monads for the Working Lisp Programmer}
% \subtitle{Subtitle Text, if any}

\authorinfo{Ravi Nanavati\and Jeff Newbern}
           {Bluespec, Inc.}
           {\{ravi,jnewbern\}@bluespec.com}

\maketitle

\begin{abstract}

Monads are a computational pattern for encapsulating and controlling
effects.  First employed in pure languages, monad design patterns have
recently been implemented in impure functional languages.  This
tutorial introduces monads as implemented in the pure language Haskell
before describing monad facilities for Scheme and Clojure.  An extended
example of monadic programming in Clojure is developed by building a
modular language interpreter.

\end{abstract}

\category{D.1.1}{Programming Techniques}{Applicative (Functional) Programming}

\terms
Languages

\keywords
Functional Programming, Haskell, Lisp, Scheme, Clojure, Monads, Monad Transformers, Modular Interpreters 

\section{Introduction}

The monad design pattern adds structure to a value so that effects
associated with computing the value can be modeled.  A wide variety of
effects can be modeled this way, including things that are not
normally thought of as effects, such as parser failure or accessing an
environment of bindings
\citep{moggi-monads,wadlermonads, parser-combinators}.

For the programmer, monads are useful tools for structuring functional
programs because they provide modularity, flexibility and isolation
\citep{all-about-monads, pureshirt}.
Modularity is achieved because monads allow effectful computations to
be composed from simpler effectful computations. Monadic programs can be
more flexible than non-monadic programs because monads separate
effectful processing and the plumbing required to implement effects --
allowing both plumbing and processing to be changed or replaced
independently. Monads also make it easier to isolate effectful
computations and to track and delimit the scope of effects within a
program.

\subsection{Definition}

There are three components to a computational monad:
\begin{enumerate}
\item
A common structure around base values which will model the
effect of the monad.  We can represent the type of an arbitrary
base value by $t$. Corresponding to $t$, the type of a monadic
value which extends the base value with a model of the desired
effect can be described by writing the name of the effect
in a subscript, $t_{effect}$.

\item
A polymorphic function, \texttt{result}, of type $t \rightarrow t_{effect}$
which takes a value of the base type $t$ and lifts it to a value of the
monadic type by adding structure corresponding to a pure computation
with no effect.

\item
A second polymorphic function, \texttt{bind}, of type
$t_{effect} \rightarrow (t \rightarrow t'_{effect}) \rightarrow t'_{effect}$.
\texttt{bind} combines a monadic value with a function that produces a new 
monadic value. The value of type $t'_{effect}$ in the result of \texttt{bind} 
is computed from the input value of type $t$ by applying the input function.  
The effect structure of the result of \texttt{bind} is the composition of 
the effect structure of the input value with the effect structure of the 
result of the function application. In other words, \texttt{bind} composes 
effectful computations.
\end{enumerate}

Since the \texttt{bind} function only composes existing effect structures
and the \texttt{result} function lifts a value purely -- that is without
adding an effect -- any useful monad (other than the identity monad)
will be augmented with additional functions which create monadic
values with effects.  Unlike \texttt{result} and \texttt{bind}, the names
and types of these functions are not standardized and will vary
between different monads.

\subsection{Monad Laws}

The \texttt{result} and \texttt{bind} functions together describe a system
for composing computations with effects.  To ensure that a monad
implements a consistent effect system, the monad should obey three laws:

\begin{enumerate}
\item \texttt{bind (result x) f $\equiv$ f x}
\item \texttt{bind m result $\equiv$ m}
\item \texttt{bind (bind m f) g $\equiv$ bind m ($\lambda$x $\rightarrow$ bind (f x) g)}
\end{enumerate}

The first law requires that \texttt{result} is a left-identity with
respect to \texttt{bind}. The second law requires that \texttt{result} is a
right-identity with respect to \texttt{bind}. The third law is a kind of
associativity law for \texttt{bind}.

\subsection{Monad Transformers}

Monad transformers allow monads for different effects to be composed
into a combined monad.  Multiple monad transformers can be applied to
a base monad to create a ``transformer stack'' that combines the
effects of each individual monad in the stack.  In general monad
composition is not commutative, so the order in which transformers are
applied in the stack affects the semantics of the combined monad.

The identity monad does not add any effect structure to the base value
-- its \texttt{return} function is the identity and its \texttt{bind} is
ordinary function application.  Any monad and its monad transformer
are related through the identity monad, since the monad is equivalent
to applying the monad transformer to the identity monad.

\section{Monads in Haskell}

\newcommand{\hbind}[0]{\texttt{>>=}}

In the pure, non-strict, statically-typed functional language Haskell
monads are pervasive \citep{awkward-squad,haskell98}.  Impure features
such as I/O and mutable state are built on a monadic framework and
monads are used to model failure, nondeterminism, partiality, parsing,
and numerous other effects.

In Haskell a monad is represented by a constructor for the type of
monadic values given the type of the base value, a {\tt return}
function for lifting values of the base type to the monadic type, and
an infix binding combinator, \hbind, for composing a value of the
monadic type with a function to create a new value of the monadic
type, combining the effects of the input value and the function.

\begin{figure}
\begin{center}
\begin{verbatim}
class Monad m where
    (>>=)  :: m a -> (a -> m b) -> m b
    return :: a -> m a
\end{verbatim}
\end{center}
\caption{Haskell's Monad class}
\label{fig-Haskell-Monad-class}
\end{figure}

Haskell's {\tt return} function represents \texttt{result} and
\hbind \linebreak represents \texttt{bind}.

Haskell includes a special syntax, {\tt do}-notation, that provides
an imperative interface for describing monadic computations. This 
interface is easier to use than manually applying \hbind \ and {\tt return}
to structure large monadic computations. Haskell also provides a
library of standard monads and monad transformers, including
transformers for working with state, continuations, failure,
environments, logging and errors.  In addition, users can easily
define their own monads and monad transformers. Both user and
library monads work seamlessly with the {\tt do}-notation and
the generic monad functions provided by standard Haskell 
libraries.

\section{Monads in Scheme}

Scheme does not have native support for monads, but, using macros,
a number of different monad packages have been developed. Unlike 
Clojure, no monad package has become a de-facto
standard for Scheme. However, one Scheme monad package \citep{scheme-monads} 
is particularly simple and provides a good framework for discussing 
the implementation of monadic syntax. In the full tutorial, this monad 
package and the implementation of its {\tt letM} and {\tt letM*} macros
will be developed in order to demonstrate how a Lisp programmer can easily 
build their own monad library.

\section{Monads in Clojure}

Clojure has a growing library of contributed code for working with
monads. This library, {\tt clojure.contrib.monads}, is patterned after 
Haskell's monad libraries \citep{clojure-monads1, clojure-monads2}.

Monads are represented by structures with standard field names
{\tt m-result} and {\tt m-bind}. The library includes a number of
macros that provide convenient monadic syntax defined in terms
of accesses to these fields of a monad structure.

The {\tt with-monad} macro takes a monad structure and a number of
expressions as arguments and binds the {\tt m-bind} and {\tt
  m-result} fields of the monad structure to local names so that any
occurence of {\tt m-bind} or {\tt m-result} in the expressions
will refer to the fields of the selected monad.  The expressions are
said to be ``run in'' the selected monad.  Often monadic code can only
be run in a single monad, but sometimes it is possible to write code
which can run in different monads to achieve different effects
(eg. running in either an interpretation monad or a compilation
monad).  This definition:
\begin{verbatim}
(defn lookup [key]
  (with-monad maybe
    (let [entry (find db key)]
       (if entry
           (m-result (second entry))
           m-zero))))
\end{verbatim}
shows a simple example of {\tt with-monad} used to run a computation
in the {\tt maybe} monad.

The {\tt domonad} macro is used for monad comprehensions with a list
of monadic bindings and a final expression.  Each of the bindings is
evaluated in turn, making the value accessible to all later bindings
and the final expression.  The final result of the {\tt domonad}
comprehension is the value of the final expression returned using
the monad's {\tt m-result} function.

For example, the comprehension:
\begin{verbatim}
(domonad maybe
  ; bindings
  [record (lookup employee)]
  ; final expression
  (:last-name record))
\end{verbatim}
is equivalent to the unsugared form:
\begin{verbatim}
(bind (lookup employee)
      (fn [record]
        (m-result (:last-name record))))
\end{verbatim}

The {\tt defmonad} macro can be used to create user-defined named
monads.  It is also used internally to define the monads supplied
with the library, such as the {\tt maybe} monad 
(see Figure \ref{fig-Clojure-maybe-defn}).

\begin{figure}
\begin{center}
\begin{verbatim}
(defmonad maybe
   "Monad describing computations with possible
    failures. Failure is represented by nil,
    any other value is considered valid.
    As soon as a step returns nil, the whole
    computation will yield nil as well."
   [m-zero   nil
    m-result (fn m-result-maybe [v] v)
    m-bind   (fn m-bind-maybe [mv f]
               (if (nil? mv) nil (f mv)))
    m-plus   (fn m-plus-maybe [& mvs]
               (first (drop-while nil? mvs)))
    ])

\end{verbatim}
\end{center}
\caption{Clojure's maybe monad definition}
\label{fig-Clojure-maybe-defn}
\end{figure}

\section{A Modular Interpreter}

A simple interpreter provides a nice example of modular monad
composition in Clojure.  Different effects in the interpreter are each
encapsulated in different monads, providing a set of language modules
which can be composed using different monad transformer combinations
to build interpreters for different language subsets
\citep{modular-interpreters, semantic-lego}.

\subsection {A Simple Arithmetic Language}

The first fragment of our interpreter implements a simple 
arithmetic expression language supporting addition, subtraction, 
multiplication and division. Division is a partial function 
since there is no result when the denominator is zero.  The 
implementation uses an error monad to support partiality as an effect.  
The error monad distinguishes between defined values and undefined values, 
which are associated with a failure description.

\begin{figure}
\begin{center}
\begin{verbatim}
(defn interp [e]
  (cond
   (number? e) 
     (domonad interp-monad [] e)
   (seq? e)    
     (let [t (first e)
           args (rest e)]
       (cond
         (= t '+) 
           (domonad interp-monad
              [x (interp (first args))
               y (interp (second args))]
             (+ x y))
         (= t '-) 
            (domonad interp-monad
              [x (interp (first args))
               y (interp (second args))]
             (- x y))
         (= t '*) 
            (domonad interp-monad
              [x (interp (first args))
               y (interp (second args))]
             (* x y))
         (= t '/) 
            (domonad interp-monad
              [x (interp (first args))
               y (interp (second args))
               r (if (and (number? y) 
                          (not= 0 y))
                     (m-result (/ x y))
                     (report-error 
                        "division by 0"))]
             r)))))
\end{verbatim}
\end{center}
\caption{Interpreter for an arithmetic expression language}
\label{fig-interp-0}
\end{figure}

The interpreter in Figure \ref{fig-interp-0} can be run in the error
monad to evaluate simple arithmetic expressions.  For {\tt '(+ 2 2)}
the interpreter returns {\tt (ok 4)}, but {\tt '(/ 12 0)} results in
{\tt (fail ``division by 0'')}.

\subsection{Adding an Environment}

To extend the interpreter with an environment of bindings an
environment monad transformer is added to the monad stack.  The
combined {\tt (env-t error)} monad provides both partiality and
binding environment effects. Adding a clause to the interpreter
to handle symbol lookup (as in Figure \ref{fig-interp-1}), 
allows the interpreter to be run with an environment of immutable 
bindings. 

\begin{figure}
\begin{center} 
\begin{verbatim}
(symbol? e)
  (domonad interp-monad
     [v (interp-lookup e)
      r (if v (interp v)
              (report-error (undef-msg e)))]
     r)
\end{verbatim}
\end{center}
\caption{Adding symbol lookup to the interpreter}
\label{fig-interp-1}
\end{figure}

\pagebreak

For example,
\begin{verbatim}
(run-with-env {'pi 3.14159} (interp '(* 2 pi)))
\end{verbatim}
yields 
\begin{verbatim}
(ok 6.28318)
\end{verbatim}
Symbol lookup can also make use of the underlying error monad
to report undefined symbols:
\begin{verbatim}
(run-with-env {'pi 3.14159} (interp '(+ 4 z)))
\end{verbatim}
returns
\begin{verbatim}
(fail "undefined variable z")
\end{verbatim}

The same binding environment can also be used to implement local
definitions and argument bindings in function abstractions, extending
the arithmetic expression language with basic means of abstraction.

\subsection{Additional Extensions}

Additional facilities can be added to the interpreter by adding
additional monads to the transformer stack. For example, adding a
state monad to the transformer stack allows the language to be
extended with a language fragment for creating, reading and writing
mutable reference cells. Including the continuation monad allows the
implementation of {\tt callcc} in the interpreter.  The ordering of
the continuation monad transformer and the state monad transformer
within the transformer stack determines whether or not calling a
continuation captured with {\tt callcc} resets or maintains the
state associated with the state monad. Tracing of interpreter 
execution can optionally be supported by incorporating a logging
monad into the transformer stack.

The use of monad transformers allows each language fragment to be
written in a modular style.  The various language building blocks can
be mixed and matched and the transformer stack can be reordered to
construct language interpreters with different feature sets and
semantics.

These examples are covered in greater detail in the full tutorial.

\section{Summary}

The full tutorial is intended to teach experienced Lisp programmers
how to work with monads.  It will begin by presenting basic monadic
concepts.  Monads in Haskell will be used to illustrate these concepts
and to introduce monad transformers.  A simple Scheme monad package
will be presented to show how a Lisp programmer can use macros to
implement their own monadic syntax.  The tutorial will conclude with
an extended example of monadic programming -- building a modular
interpreter using Clojure's monad library.

%\appendix
%\section{Appendix Title}

%This is the text of the appendix, if you need one.

%\acks

%Acknowledgments, if needed.

\bibliographystyle{plainnat}
\bibliography{ilc09-monad-tutorial}

\end{document}
